Preliminary insights into the use of pottery and culinary practices at Guijiabao site in southwest China

### 2024-04-16 analysis/paper/paper.Rmd
edit the discussion section

<!--Plant as possible non-ruminant fat origin-->
The non-ruminant fat (SYG-TN13-E23-3) could originate from C~3~ plants or freshwater resources, based on similar isotopic ranges reported for Neolithic pottery in eastern China [@Shoda2018; @Lyu2024], and a global database [@Steele2010]. However, although fish bones and net sinkers were found at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absence in our samples. This excludes the possibility of using fish fat to explain our sample. Possible source C~3~ plants could be rice or wild weeds based on botanical evidence [@Guijiabao2016; @Hao2022; @Huan2022]. Due to a small portion of seeds of wild weeds in the floral assemblage [@Guijiabao2016], coupled with rice phytolichs identified from the residential area [@Huan2022], rice appears to be the more likely major source of C~3~ plants. In addition, the non-ruminant specimen has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, indicating a possibility of plant oils [@Mayyas2018; @Copley2005]. However, we acknowledge that C~18:1~ might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have weak preservation in archaeological contexts [@Whelton2021]. A further detection of biomarkers for starchy plants with a specific derivatization method for lipid compounds will be necessary to assess the presence of rice oil [@Shoda2018].

@@

The two historical potsherds (ca. 1200-500 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C~3~ plants, pigs, or freshwater resources compared to the Neolithic period. In addition to rice, the source of C~3~ plants could also have included wheat, which was introduced into the region during the late Bronze Age (ca. 2,800–2000 cal. BP) [@Huan2022]. It is surprising that ruminant adipose fat, rather than non-ruminant fat, was detected in the Gaoshan potsherd, given the predominance of domestic pigs in the faunal assemblage at the site [@He2020]. One possibility might be our Gaoshan potsherd is an outlier, potentially not representative of the broader dietary patterns. Another plausible explanation could be that the isotopic values of ruminant adipose fat result from a mixture of non-ruminant adipose fats and ruminant milk fats. This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002].

While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels. As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the southwest China.

>>>>

<!--Plant as possible non-ruminant fat origin-->
The non-ruminant fat (SYG-TN13-E23-3) could originate from C~3~ plants or freshwater resources, based on similar isotopic ranges reported for Neolithic pottery in eastern China [@Shoda2018; @Lyu2024], and a global database [@Steele2010]. Possible source of C~3~ plants could include rice and wild weeds, supported by the presence of rice phytolichs and a small portion of carbonized seeds of wild weeds, such as *Pennisetum*, identified from the residential area at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. A further detection of biomarkers for starchy plants with a specific derivatization method for lipid compounds will be necessary to assess the presence of rice oil [@Shoda2018]. Although freshwater resources might be another source of non-ruminant fats according to the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absent in our samples. The absence of these molecules could result from either their actual absence, or poor preservation of those compounds in archaeological contexts in general. Thus, we are unable to exclude the possibility of freshwater resources.

@@

The two historical potsherds (ca. 1200-500 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C~3~ plants, pigs, or freshwater resources compared to the Neolithic period. In addition to rice, the source of C~3~ plants could also have included wheat, which was introduced into the region during the late Bronze Age (ca. 2,800–2000 cal. BP) [@Huan2022]. It is surprising that ruminant adipose fat, rather than non-ruminant fat, was detected in the Gaoshan potsherd, given the predominance of domestic pigs in the faunal assemblage at the site [@He2020]. One possibility might be our Gaoshan potsherd is an outlier, potentially not representative of the broader dietary patterns. Another plausible explanation could be that the isotopic values of ruminant adipose fat result from a mixture of non-ruminant adipose fats and ruminant milk fats. This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002]. While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels. As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the southwest China.
